---
title: "California Enrollment Deep Dive: District Trends & Demographic Shifts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{California Enrollment Deep Dive: District Trends & Demographic Shifts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  cache = TRUE
)
```

## Overview

California's public school system serves over 5.8 million students across 58 counties and
more than 1,000 school districts. This vignette explores enrollment patterns across
districts, years, and demographic subgroups to surface the most significant trends
shaping California education.

We analyze enrollment snapshots from 2018 to 2025, a period that spans pre-pandemic
normalcy, the COVID-19 disruption, and the post-pandemic recovery.

## Load Data

```{r load-libraries}
library(caschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r fetch-data, cache=FALSE}
# Fetch recent years (limited to reduce memory for vignette builds)
years <- c(2018, 2020, 2022, 2024, 2025)
enr <- fetch_enr_multi(years, use_cache = TRUE)

# Quick overview
# NOTE: charter_status == "All" is required to avoid tripled rows in 2024-2025
# (modern data reports All, Y, N separately)
enr %>%
  filter(is_state, grade_level == "TOTAL", reporting_category == "TA",
         charter_status == "All") %>%
  select(end_year, n_students) %>%
  mutate(change = n_students - lag(n_students))
```

---

## 1. California Lost Over 400,000 Students Since 2018

The most striking finding: California public schools have lost over 400,000 students
since 2018. The decline accelerated during the pandemic and has continued every year since.

```{r finding-1}
state_trend <- enr %>%
  filter(is_state, grade_level == "TOTAL", reporting_category == "TA",
         charter_status == "All") %>%
  arrange(end_year) %>%
  mutate(
    cumulative_change = n_students - first(n_students),
    pct_change = (n_students - first(n_students)) / first(n_students) * 100
  )

stopifnot(nrow(state_trend) > 0)

# Calculate the specific decline from 2018
enr_2018 <- state_trend %>% filter(end_year == 2018) %>% pull(n_students)
current <- state_trend %>% filter(end_year == max(end_year)) %>% pull(n_students)
decline <- enr_2018 - current

cat(sprintf("2018 enrollment: %s students\n", scales::comma(enr_2018)))
cat(sprintf("Current enrollment: %s students\n", scales::comma(current)))
cat(sprintf("Total decline: %s students (%.1f%%)\n",
            scales::comma(decline),
            decline / enr_2018 * 100))

ggplot(state_trend, aes(x = end_year, y = n_students)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "red", alpha = 0.7) +
  annotate("text", x = 2020.1, y = max(state_trend$n_students) * 0.98,
           label = "COVID-19", hjust = 0, color = "red") +
  scale_y_continuous(labels = scales::comma, limits = c(5500000, NA)) +
  labs(
    title = "California K-12 Public School Enrollment",
    subtitle = "2017-18 through 2024-25 School Years",
    x = "School Year (End Year)",
    y = "Total Enrollment"
  ) +
  theme_minimal()
```

---

## 2. LAUSD Has Lost the Equivalent of a Major City's School District

Los Angeles Unified, the nation's second-largest school district, has experienced
dramatic enrollment losses:

```{r finding-2}
lausd <- enr %>%
  filter(
    is_district,
    grade_level == "TOTAL",
    reporting_category == "TA",
    charter_status == "All",
    grepl("Los Angeles Unified", district_name, ignore.case = TRUE)
  ) %>%
  arrange(end_year) %>%
  mutate(
    change = n_students - lag(n_students),
    cumulative_change = n_students - first(n_students)
  )

stopifnot(nrow(lausd) > 0)

cat(sprintf("LAUSD 2018: %s students\n", scales::comma(lausd$n_students[1])))
cat(sprintf("LAUSD 2025: %s students\n", scales::comma(tail(lausd$n_students, 1))))
cat(sprintf("Total loss: %s students (%.1f%%)\n",
            scales::comma(abs(tail(lausd$cumulative_change, 1))),
            abs(tail(lausd$cumulative_change, 1)) / lausd$n_students[1] * 100))

ggplot(lausd, aes(x = end_year, y = n_students)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = scales::comma(n_students)), vjust = -0.5, size = 3) +
  scale_y_continuous(labels = scales::comma, limits = c(0, max(lausd$n_students) * 1.1)) +
  labs(
    title = "Los Angeles Unified School District Enrollment",
    subtitle = "Lost more students than the entire Fresno Unified enrollment",
    x = "School Year (End Year)",
    y = "Enrollment"
  ) +
  theme_minimal()
```

---

## 3. The Top 5 Districts Lost Over 100,000 Students Combined

California's five largest districts all face significant enrollment challenges:

```{r finding-3}
# Find the 5 largest districts (by 2025 enrollment)
top5_districts <- enr %>%
  filter(
    is_district,
    end_year == max(end_year),
    grade_level == "TOTAL",
    reporting_category == "TA",
    charter_status == "All"
  ) %>%
  arrange(desc(n_students)) %>%
  head(5) %>%
  pull(district_name)

top5_trend <- enr %>%
  filter(
    is_district,
    grade_level == "TOTAL",
    reporting_category == "TA",
    charter_status == "All",
    district_name %in% top5_districts
  ) %>%
  arrange(district_name, end_year)

stopifnot(nrow(top5_trend) > 0)

# Calculate change from first to last year
top5_change <- top5_trend %>%
  group_by(district_name) %>%
  summarize(
    enr_first = first(n_students),
    enr_last = last(n_students),
    change = last(n_students) - first(n_students),
    pct_change = (last(n_students) - first(n_students)) / first(n_students) * 100,
    .groups = "drop"
  ) %>%
  arrange(change)

top5_change %>%
  mutate(
    district_name = gsub(" School District$| Unified$| Unified School District$", "", district_name),
    change_fmt = scales::comma(change),
    pct_fmt = sprintf("%.1f%%", pct_change)
  ) %>%
  select(district_name, enr_first, enr_last, change_fmt, pct_fmt)

# Visualization
ggplot(top5_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Enrollment Trends in California's Largest Districts",
    x = "School Year (End Year)",
    y = "Enrollment",
    color = "District"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## 4. Hispanic Students Now Comprise 56% of California's Enrollment

California's demographic makeup has shifted significantly:

```{r finding-4}
# Calculate race/ethnicity percentages by year
race_by_year <- enr %>%
  filter(
    is_state,
    grade_level == "TOTAL",
    charter_status == "All",
    grepl("^RE_", reporting_category)
  ) %>%
  group_by(end_year) %>%
  mutate(
    total = sum(n_students, na.rm = TRUE),
    pct = n_students / total * 100
  ) %>%
  ungroup()

stopifnot(nrow(race_by_year) > 0)

# Latest year breakdown
latest_race <- race_by_year %>%
  filter(end_year == max(end_year)) %>%
  arrange(desc(pct)) %>%
  select(subgroup, n_students, pct) %>%
  mutate(pct_fmt = sprintf("%.1f%%", pct))

latest_race

# Visualize
ggplot(latest_race, aes(x = reorder(subgroup, pct), y = pct, fill = subgroup)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", pct)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 65)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "California Enrollment by Race/Ethnicity",
    subtitle = sprintf("School Year %d-%d", max(enr$end_year) - 1, max(enr$end_year)),
    x = NULL,
    y = "Percent of Students"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## 5. Some Districts Grew While Others Collapsed

Not all districts experienced decline. A handful of districts bucked the statewide
trend with substantial growth:

```{r finding-5}
# Calculate district change from 2020 to latest year
district_changes <- enr %>%
  filter(
    is_district,
    grade_level == "TOTAL",
    reporting_category == "TA",
    charter_status == "All",
    end_year %in% c(2020, max(end_year))
  ) %>%
  pivot_wider(
    id_cols = c(district_name, county_name, cds_code),
    names_from = end_year,
    values_from = n_students,
    names_prefix = "enr_"
  ) %>%
  filter(!is.na(enr_2020) & enr_2020 > 1000) %>%  # Filter to districts with baseline data
  mutate(
    change = .[[ncol(.)]] - enr_2020,
    pct_change = change / enr_2020 * 100
  )

stopifnot(nrow(district_changes) > 0)

# Top 10 growing districts
top_growers <- district_changes %>%
  arrange(desc(pct_change)) %>%
  head(10) %>%
  select(district_name, county_name, enr_2020, change, pct_change) %>%
  mutate(
    enr_2020 = scales::comma(enr_2020),
    change = paste0("+", scales::comma(change)),
    pct_change = sprintf("+%.1f%%", pct_change)
  )

cat("Top 10 Growing Districts (2020 to Present):\n\n")
print(top_growers, n = 10)

# Top 10 declining districts (by percentage)
top_decliners <- district_changes %>%
  filter(enr_2020 > 5000) %>%  # Only larger districts
  arrange(pct_change) %>%
  head(10) %>%
  select(district_name, county_name, enr_2020, change, pct_change) %>%
  mutate(
    enr_2020 = scales::comma(enr_2020),
    change = scales::comma(change),
    pct_change = sprintf("%.1f%%", pct_change)
  )

cat("\nTop 10 Declining Districts (2020 to Present, Districts >5,000 students):\n\n")
print(top_decliners, n = 10)
```

---

## 6. High School Enrollment Dropped Faster Than Elementary

Enrollment loss varied significantly by grade level:

```{r finding-6}
# Grade-level trends (state level)
grade_trends <- enr %>%
  filter(
    is_state,
    reporting_category == "TA",
    charter_status == "All",
    grade_level %in% c("K", "01", "02", "03", "04", "05",
                        "06", "07", "08", "09", "10", "11", "12")
  ) %>%
  mutate(
    grade_band = case_when(
      grade_level %in% c("K", "01", "02", "03", "04", "05") ~ "Elementary (K-5)",
      grade_level %in% c("06", "07", "08") ~ "Middle (6-8)",
      TRUE ~ "High (9-12)"
    )
  ) %>%
  group_by(end_year, grade_band) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

stopifnot(nrow(grade_trends) > 0)

# Calculate change from first year
grade_change <- grade_trends %>%
  group_by(grade_band) %>%
  mutate(
    pct_of_first = n_students / first(n_students) * 100,
    index = n_students / first(n_students) * 100
  ) %>%
  ungroup()

ggplot(grade_change, aes(x = end_year, y = index, color = grade_band)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 100, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = c("Elementary (K-5)" = "#1b9e77",
                                 "Middle (6-8)" = "#d95f02",
                                 "High (9-12)" = "#7570b3")) +
  labs(
    title = "Enrollment Change by Grade Band (Indexed to 2018 = 100)",
    subtitle = "All grade bands have declined, but high school dropped earliest",
    x = "School Year (End Year)",
    y = "Enrollment Index (2018 = 100)",
    color = "Grade Band"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## 7. The Bay Area Exodus: Tech Counties Lost the Most Students

San Francisco, Santa Clara, and other Bay Area counties experienced some of the
steepest enrollment drops:

```{r finding-7}
county_changes <- enr %>%
  filter(
    is_county,
    grade_level == "TOTAL",
    reporting_category == "TA",
    charter_status == "All",
    end_year %in% c(2020, max(end_year))
  ) %>%
  pivot_wider(
    id_cols = c(county_name),
    names_from = end_year,
    values_from = n_students,
    names_prefix = "enr_"
  ) %>%
  filter(!is.na(enr_2020)) %>%
  mutate(
    change = .[[ncol(.)]] - enr_2020,
    pct_change = change / enr_2020 * 100
  ) %>%
  arrange(pct_change)

stopifnot(nrow(county_changes) > 0)

# Top 10 counties with biggest percentage decline
cat("Top 10 Counties with Largest Enrollment Decline (2020 to Present):\n\n")
county_changes %>%
  head(10) %>%
  select(county_name, enr_2020, change, pct_change) %>%
  mutate(
    enr_2020 = scales::comma(enr_2020),
    change = scales::comma(change),
    pct_change = sprintf("%.1f%%", pct_change)
  )

# Visualize county changes
county_plot_data <- county_changes %>%
  mutate(
    region = case_when(
      county_name %in% c("San Francisco", "Santa Clara", "Alameda",
                          "San Mateo", "Contra Costa", "Marin") ~ "Bay Area",
      county_name %in% c("Los Angeles", "Orange", "San Diego",
                          "Riverside", "San Bernardino") ~ "SoCal Metro",
      TRUE ~ "Other"
    )
  )

ggplot(county_plot_data, aes(x = pct_change, fill = region)) +
  geom_histogram(bins = 20, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("Bay Area" = "#e41a1c",
                                "SoCal Metro" = "#377eb8",
                                "Other" = "#999999")) +
  labs(
    title = "Distribution of County Enrollment Changes (2020 to Present)",
    subtitle = "Bay Area counties cluster at the far left (biggest losses)",
    x = "Percent Change in Enrollment",
    y = "Number of Counties",
    fill = "Region"
  ) +
  theme_minimal()
```

---

## 8. Kindergarten Enrollment Signals Future Decline

Kindergarten enrollment is a leading indicator for future enrollment. The drop in
K enrollment since 2020 suggests continued overall declines ahead:

```{r finding-8}
k_trend <- enr %>%
  filter(
    is_state,
    reporting_category == "TA",
    charter_status == "All",
    grade_level == "K"
  ) %>%
  arrange(end_year) %>%
  mutate(
    change = n_students - lag(n_students),
    pct_change = (n_students - lag(n_students)) / lag(n_students) * 100
  )

stopifnot(nrow(k_trend) > 0)

cat(sprintf("Kindergarten Enrollment 2018: %s\n", scales::comma(k_trend$n_students[1])))
cat(sprintf("Kindergarten Enrollment %d: %s\n", max(k_trend$end_year),
            scales::comma(tail(k_trend$n_students, 1))))
cat(sprintf("Change: %s (%.1f%%)\n",
            scales::comma(tail(k_trend$n_students, 1) - k_trend$n_students[1]),
            (tail(k_trend$n_students, 1) - k_trend$n_students[1]) / k_trend$n_students[1] * 100))

ggplot(k_trend, aes(x = end_year, y = n_students)) +
  geom_line(color = "#d95f02", linewidth = 1.2) +
  geom_point(color = "#d95f02", size = 3) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "California Kindergarten Enrollment",
    subtitle = "Kindergarten trends forecast future total enrollment",
    x = "School Year (End Year)",
    y = "Kindergarten Students"
  ) +
  theme_minimal()
```

---

## 9. Gender Ratios Have Remained Remarkably Stable

Despite major enrollment shifts, the gender ratio has stayed nearly constant:

```{r finding-9}
gender_trend <- enr %>%
  filter(
    is_state,
    grade_level == "TOTAL",
    charter_status == "All",
    reporting_category %in% c("GN_F", "GN_M")
  ) %>%
  group_by(end_year) %>%
  mutate(
    total = sum(n_students),
    pct = n_students / total * 100
  ) %>%
  ungroup()

stopifnot(nrow(gender_trend) > 0)

gender_wide <- gender_trend %>%
  select(end_year, subgroup, pct) %>%
  pivot_wider(names_from = subgroup, values_from = pct)

gender_wide

ggplot(gender_trend, aes(x = end_year, y = pct, fill = subgroup)) +
  geom_col(position = "stack") +
  geom_hline(yintercept = 50, linetype = "dashed", color = "white", linewidth = 1) +
  scale_fill_manual(values = c("female" = "#e78ac3", "male" = "#66c2a5")) +
  labs(
    title = "Gender Distribution Over Time",
    subtitle = "Male/Female ratio has remained stable at roughly 51/49",
    x = "School Year (End Year)",
    y = "Percent of Students",
    fill = "Gender"
  ) +
  theme_minimal()
```

---

## 10. English Learner Population Remains Substantial

English Learners represent a significant and consistent portion of California's
student population (data available for 2024-2025):

```{r finding-10}
el_data <- enr %>%
  filter(
    is_state,
    grade_level == "TOTAL",
    charter_status == "All",
    reporting_category %in% c("TA", "SG_EL"),
    end_year >= 2024  # SG_EL only available in modern data
  ) %>%
  pivot_wider(
    id_cols = end_year,
    names_from = reporting_category,
    values_from = n_students
  ) %>%
  mutate(
    el_pct = SG_EL / TA * 100
  )

if (nrow(el_data) > 0) {
  cat("English Learner Enrollment:\n")
  el_data %>%
    mutate(
      total = scales::comma(TA),
      el = scales::comma(SG_EL),
      el_pct = sprintf("%.1f%%", el_pct)
    ) %>%
    select(end_year, total, el, el_pct)
}

# Show student group breakdown for latest year
student_groups <- enr %>%
  filter(
    is_state,
    grade_level == "TOTAL",
    charter_status == "All",
    grepl("^SG_", reporting_category),
    end_year == max(end_year)
  ) %>%
  arrange(desc(n_students)) %>%
  select(subgroup, n_students)

if (nrow(student_groups) > 0) {
  cat("\nStudent Group Populations (Latest Year):\n")
  student_groups %>%
    mutate(n_students = scales::comma(n_students))
}
```

---

## Summary: Key Findings

| # | Finding | Key Metric |
|---|---------|------------|
| 1 | California lost over 400,000 students since 2018 | ~6.7% decline statewide |
| 2 | LAUSD lost the equivalent of a major city's district | ~80,000+ students |
| 3 | Top 5 districts lost 100,000+ students combined | Double-digit % declines |
| 4 | Hispanic students now 56% of enrollment | Up from historical levels |
| 5 | Some districts grew while most declined | Wide variation in trends |
| 6 | High school enrollment dropped fastest | Different pipeline impacts |
| 7 | Bay Area counties lost the most students (%) | Tech region exodus |
| 8 | Kindergarten decline signals future drops | Leading indicator |
| 9 | Gender ratios remained stable | ~51% male, ~49% female |
| 10 | English Learners remain a major population | 18%+ of students |

---

## Data Notes

- **Years 2018-2023**: Historical school-level files with race/ethnicity and gender data.
  District, county, and state aggregates computed from school records. Charter status = "All".
- **Years 2024-2025**: Modern Census Day files with complete demographic breakdowns.
  including student group categories (English Learners, Students with Disabilities, etc.)
- **Suppression**: Counts of 10 or fewer students are suppressed for privacy
- **Census Day**: All enrollment counts are from Census Day (first Wednesday in October)

## Session Info

```{r session-info}
sessionInfo()
```
