---
title: "California Enrollment: District Trends & Demographics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{California Enrollment: District Trends & Demographics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

## Overview

California's public school system serves over 5.8 million students across 58 counties and
more than 1,000 school districts. This vignette explores enrollment patterns across
districts, years, and demographic subgroups.

We analyze data from 2023-2025 (school years 2022-23 through 2024-25).

## Load Data

```{r load-libraries}
library(caschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r fetch-data}
# Fetch recent years (2023-2025)
years <- 2023:2025
enr <- fetch_enr_multi(years, use_cache = TRUE)

# Quick overview
state_totals <- enr %>%
  filter(is_state, grade_level == "TOTAL", reporting_category == "TA") %>%
  select(end_year, n_students) %>%
  mutate(change = n_students - lag(n_students))

state_totals
```

---

## 1. State Enrollment Trend

```{r state-trend}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  scale_y_continuous(labels = scales::comma, limits = c(5500000, NA)) +
  labs(
    title = "California K-12 Public School Enrollment",
    subtitle = "2022-23 through 2024-25 School Years",
    x = "School Year (End Year)",
    y = "Total Enrollment"
  ) +
  theme_minimal()
```

---

## 2. LAUSD Enrollment

Los Angeles Unified is the nation's second-largest school district:

```{r lausd}
lausd <- enr %>%
  filter(
    is_district,
    grade_level == "TOTAL",
    reporting_category == "TA",
    grepl("Los Angeles Unified", district_name, ignore.case = TRUE)
  ) %>%
  arrange(end_year) %>%
  mutate(
    change = n_students - lag(n_students),
    cumulative_change = n_students - first(n_students)
  )

cat(sprintf("LAUSD %d: %s students\n", min(lausd$end_year), scales::comma(lausd$n_students[1])))
cat(sprintf("LAUSD %d: %s students\n", max(lausd$end_year), scales::comma(tail(lausd$n_students, 1))))
cat(sprintf("Change: %s students (%.1f%%)\n",
            scales::comma(tail(lausd$cumulative_change, 1)),
            tail(lausd$cumulative_change, 1) / lausd$n_students[1] * 100))

ggplot(lausd, aes(x = end_year, y = n_students)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = scales::comma(n_students)), vjust = -0.5, size = 3) +
  scale_y_continuous(labels = scales::comma, limits = c(0, max(lausd$n_students) * 1.1)) +
  labs(
    title = "Los Angeles Unified School District Enrollment",
    x = "School Year (End Year)",
    y = "Enrollment"
  ) +
  theme_minimal()
```

---

## 3. Top 5 Districts

California's five largest districts:

```{r top5}
# Find the 5 largest districts (by latest year enrollment)
top5_districts <- enr %>%
  filter(
    is_district,
    end_year == max(end_year),
    grade_level == "TOTAL",
    reporting_category == "TA"
  ) %>%
  arrange(desc(n_students)) %>%
  head(5) %>%
  pull(district_name)

top5_trend <- enr %>%
  filter(
    is_district,
    grade_level == "TOTAL",
    reporting_category == "TA",
    district_name %in% top5_districts
  ) %>%
  arrange(district_name, end_year)

# Calculate change from first to last year
top5_change <- top5_trend %>%
  group_by(district_name) %>%
  summarize(
    enr_first = first(n_students),
    enr_last = last(n_students),
    change = last(n_students) - first(n_students),
    pct_change = (last(n_students) - first(n_students)) / first(n_students) * 100,
    .groups = "drop"
  ) %>%
  arrange(change)

top5_change %>%
  mutate(
    district_name = gsub(" School District$| Unified$| Unified School District$", "", district_name),
    change_fmt = scales::comma(change),
    pct_fmt = sprintf("%.1f%%", pct_change)
  ) %>%
  select(district_name, enr_first, enr_last, change_fmt, pct_fmt)

# Visualization
ggplot(top5_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Enrollment Trends in California's Largest Districts",
    x = "School Year (End Year)",
    y = "Enrollment",
    color = "District"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## 4. Demographics: Race/Ethnicity

California's demographic makeup:

```{r demographics}
# Latest year breakdown
latest_race <- enr %>%
  filter(
    is_state,
    grade_level == "TOTAL",
    grepl("^RE_", reporting_category),
    end_year == max(end_year)
  ) %>%
  mutate(
    total = sum(n_students, na.rm = TRUE),
    pct = n_students / total * 100
  ) %>%
  arrange(desc(pct)) %>%
  select(subgroup, n_students, pct) %>%
  mutate(pct_fmt = sprintf("%.1f%%", pct))

latest_race

# Visualize
ggplot(latest_race, aes(x = reorder(subgroup, pct), y = pct, fill = subgroup)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", pct)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 65)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "California Enrollment by Race/Ethnicity",
    subtitle = sprintf("School Year %d-%d", max(enr$end_year) - 1, max(enr$end_year)),
    x = NULL,
    y = "Percent of Students"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## 5. District Growth and Decline

Comparing district enrollment changes:

```{r district-changes}
# Calculate district change from first to latest year
min_year <- min(enr$end_year)
max_year <- max(enr$end_year)

# Get first year data
first_year_data <- enr %>%
  filter(
    is_district,
    grade_level == "TOTAL",
    reporting_category == "TA",
    end_year == min_year
  ) %>%
  select(cds_code, district_name, county_name, enr_first = n_students)

# Get last year data
last_year_data <- enr %>%
  filter(
    is_district,
    grade_level == "TOTAL",
    reporting_category == "TA",
    end_year == max_year
  ) %>%
  select(cds_code, enr_last = n_students)

# Join and calculate changes
district_changes <- first_year_data %>%
  inner_join(last_year_data, by = "cds_code") %>%
  filter(enr_first > 1000) %>%
  mutate(
    change = enr_last - enr_first,
    pct_change = change / enr_first * 100
  )

# Top 10 growing districts
top_growers <- district_changes %>%
  arrange(desc(pct_change)) %>%
  head(10) %>%
  select(district_name, county_name, enr_first, change, pct_change)

cat(sprintf("Top 10 Growing Districts (%d to %d):\n\n", min_year, max_year))
print(top_growers, n = 10)

# Top 10 declining districts (by percentage)
top_decliners <- district_changes %>%
  filter(enr_first > 5000) %>%
  arrange(pct_change) %>%
  head(10) %>%
  select(district_name, county_name, enr_first, change, pct_change)

cat(sprintf("\nTop 10 Declining Districts (%d to %d, Districts >5,000 students):\n\n",
            min_year, max_year))
print(top_decliners, n = 10)
```

---

## 6. Grade Band Analysis

Enrollment by grade level:

```{r grade-bands}
# Grade-level trends (state level)
grade_trends <- enr %>%
  filter(
    is_state,
    reporting_category == "TA",
    grade_level %in% c("K", "01", "02", "03", "04", "05",
                        "06", "07", "08", "09", "10", "11", "12")
  ) %>%
  mutate(
    grade_band = case_when(
      grade_level %in% c("K", "01", "02", "03", "04", "05") ~ "Elementary (K-5)",
      grade_level %in% c("06", "07", "08") ~ "Middle (6-8)",
      TRUE ~ "High (9-12)"
    )
  ) %>%
  group_by(end_year, grade_band) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

# Calculate change from first year
grade_change <- grade_trends %>%
  group_by(grade_band) %>%
  mutate(
    index = n_students / first(n_students) * 100
  ) %>%
  ungroup()

ggplot(grade_change, aes(x = end_year, y = index, color = grade_band)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 100, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = c("Elementary (K-5)" = "#1b9e77",
                                 "Middle (6-8)" = "#d95f02",
                                 "High (9-12)" = "#7570b3")) +
  labs(
    title = sprintf("Enrollment Change by Grade Band (Indexed to %d = 100)", min(enr$end_year)),
    x = "School Year (End Year)",
    y = sprintf("Enrollment Index (%d = 100)", min(enr$end_year)),
    color = "Grade Band"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## 7. County Analysis

Enrollment changes by county:

```{r county-analysis}
# Get first year county data
county_first <- enr %>%
  filter(
    is_county,
    grade_level == "TOTAL",
    reporting_category == "TA",
    end_year == min_year
  ) %>%
  select(county_name, enr_first = n_students)

# Get last year county data
county_last <- enr %>%
  filter(
    is_county,
    grade_level == "TOTAL",
    reporting_category == "TA",
    end_year == max_year
  ) %>%
  select(county_name, enr_last = n_students)

# Join and calculate changes
county_changes <- county_first %>%
  inner_join(county_last, by = "county_name") %>%
  mutate(
    change = enr_last - enr_first,
    pct_change = change / enr_first * 100
  ) %>%
  arrange(pct_change)

# Top 10 counties with biggest percentage decline
cat(sprintf("Top 10 Counties with Largest Enrollment Decline (%d to %d):\n\n",
            min_year, max_year))
county_changes %>%
  head(10) %>%
  select(county_name, enr_first, change, pct_change) %>%
  mutate(pct_change = sprintf("%.1f%%", pct_change))

# Visualize county changes
county_plot_data <- county_changes %>%
  mutate(
    region = case_when(
      county_name %in% c("San Francisco", "Santa Clara", "Alameda",
                          "San Mateo", "Contra Costa", "Marin") ~ "Bay Area",
      county_name %in% c("Los Angeles", "Orange", "San Diego",
                          "Riverside", "San Bernardino") ~ "SoCal Metro",
      TRUE ~ "Other"
    )
  )

ggplot(county_plot_data, aes(x = pct_change, fill = region)) +
  geom_histogram(bins = 20, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("Bay Area" = "#e41a1c",
                                "SoCal Metro" = "#377eb8",
                                "Other" = "#999999")) +
  labs(
    title = sprintf("Distribution of County Enrollment Changes (%d to %d)",
                    min(enr$end_year), max(enr$end_year)),
    x = "Percent Change in Enrollment",
    y = "Number of Counties",
    fill = "Region"
  ) +
  theme_minimal()
```

---

## 8. Kindergarten Enrollment

Kindergarten enrollment as a leading indicator:

```{r kindergarten}
k_trend <- enr %>%
  filter(
    is_state,
    reporting_category == "TA",
    grade_level == "K"
  ) %>%
  arrange(end_year) %>%
  mutate(
    change = n_students - lag(n_students),
    pct_change = (n_students - lag(n_students)) / lag(n_students) * 100
  )

cat(sprintf("Kindergarten Enrollment %d: %s\n", min(k_trend$end_year), scales::comma(k_trend$n_students[1])))
cat(sprintf("Kindergarten Enrollment %d: %s\n", max(k_trend$end_year),
            scales::comma(tail(k_trend$n_students, 1))))
cat(sprintf("Change: %s (%.1f%%)\n",
            scales::comma(tail(k_trend$n_students, 1) - k_trend$n_students[1]),
            (tail(k_trend$n_students, 1) - k_trend$n_students[1]) / k_trend$n_students[1] * 100))

ggplot(k_trend, aes(x = end_year, y = n_students)) +
  geom_line(color = "#d95f02", linewidth = 1.2) +
  geom_point(color = "#d95f02", size = 3) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "California Kindergarten Enrollment",
    subtitle = "Kindergarten trends forecast future total enrollment",
    x = "School Year (End Year)",
    y = "Kindergarten Students"
  ) +
  theme_minimal()
```

---

## 9. Gender Distribution

```{r gender}
gender_trend <- enr %>%
  filter(
    is_state,
    grade_level == "TOTAL",
    reporting_category %in% c("GN_F", "GN_M")
  ) %>%
  group_by(end_year) %>%
  mutate(
    total = sum(n_students),
    pct = n_students / total * 100
  ) %>%
  ungroup()

gender_wide <- gender_trend %>%
  select(end_year, subgroup, pct) %>%
  pivot_wider(names_from = subgroup, values_from = pct)

gender_wide

ggplot(gender_trend, aes(x = end_year, y = pct, fill = subgroup)) +
  geom_col(position = "stack") +
  geom_hline(yintercept = 50, linetype = "dashed", color = "white", linewidth = 1) +
  scale_fill_manual(values = c("female" = "#e78ac3", "male" = "#66c2a5")) +
  labs(
    title = "Gender Distribution Over Time",
    subtitle = "Male/Female ratio has remained stable at roughly 51/49",
    x = "School Year (End Year)",
    y = "Percent of Students",
    fill = "Gender"
  ) +
  theme_minimal()
```

---

## 10. Student Groups (Modern Data)

English Learners and other student groups (available for 2024-2025):

```{r student-groups}
# Get total enrollment for modern years
total_by_year <- enr %>%
  filter(
    is_state,
    grade_level == "TOTAL",
    reporting_category == "TA",
    end_year >= 2024
  ) %>%
  select(end_year, total = n_students)

# Get EL enrollment for modern years
el_by_year <- enr %>%
  filter(
    is_state,
    grade_level == "TOTAL",
    reporting_category == "SG_EL",
    end_year >= 2024
  ) %>%
  select(end_year, el = n_students)

# Join and calculate percentage
el_data <- total_by_year %>%
  left_join(el_by_year, by = "end_year") %>%
  filter(!is.na(el)) %>%
  mutate(
    el_pct = el / total * 100
  )

if (nrow(el_data) > 0) {
  cat("English Learner Enrollment:\n")
  el_data %>%
    mutate(
      total = scales::comma(total),
      el = scales::comma(el),
      el_pct = sprintf("%.1f%%", el_pct)
    ) %>%
    select(end_year, total, el, el_pct) %>%
    print()
}

# Show student group breakdown for latest year
student_groups <- enr %>%
  filter(
    is_state,
    grade_level == "TOTAL",
    grepl("^SG_", reporting_category),
    end_year == max(end_year)
  ) %>%
  arrange(desc(n_students)) %>%
  select(subgroup, n_students)

if (nrow(student_groups) > 0) {
  cat("\nStudent Group Populations (Latest Year):\n")
  student_groups %>%
    mutate(n_students = scales::comma(n_students)) %>%
    print()
}
```

---

## Data Notes

- **Years 2023**: Historical school-level files with race/ethnicity and gender data.
  District, county, and state aggregates computed from school records.
- **Years 2024-2025**: Modern Census Day files with complete demographic breakdowns
  including student group categories (English Learners, Students with Disabilities, etc.)
- **Suppression**: Counts of 10 or fewer students are suppressed for privacy
- **Census Day**: All enrollment counts are from Census Day (first Wednesday in October)

## Session Info

```{r session-info}
sessionInfo()
```
